---
title: "Grade mark correlation features for university students"
author: "Bananchiki group: Lev Laptinov, Dmytro Kolisnyk, Vasyl Moiseiev"
---

# Introduction

[Source of dataset](https://archive.ics.uci.edu/dataset/697/predict+students+dropout+and+academic+success)

A dataset created from a higher education institution (acquired from several disjoint databases) related to students enrolled in different undergraduate degrees, such as agronomy, design, education, nursing, journalism, management, social service, and technologies. The dataset includes information known at the time of student enrollment (academic path, demographics, and social-economic factors) and the students' academic performance at the end of the first and second semesters. [Dataset is already performed a rigorous data preprocessing to handle data from anomalies, unexplainable outliers, and missing values.](https://arc.net/l/quote/emkqvmzr)

```{r, include=FALSE}
library(tidyverse)
library(rvest)
library(ggplot2)
if (!require("skimr")) {
  install.packages("skimr")
}
library(skimr)
```

```{r, include=FALSE}
data <- read_delim("data.csv", delim = ";")
head(data)
```

```{r}
selected_data <- data %>%
    select(
        attendance_type = `Daytime/evening attendance\t`,
        previous_qualification_grade = `Previous qualification (grade)`,
        admission_grade = `Admission grade`,
        `Gender`,
        scholarship_holder= `Scholarship holder`,
        age_at_enrollment = `Age at enrollment`,
        first_sem_grade = `Curricular units 1st sem (grade)`,
        second_sem_grade = `Curricular units 2nd sem (grade)`
    )
head(selected_data)
```

```{r}
selected_data$first_year_grade <- (selected_data$first_sem_grade + selected_data$second_sem_grade)/2
selected_data$attendance_type <- factor(ifelse(selected_data$attendance_type == 1, "daytime", "evening"))
selected_data$Gender <- factor(ifelse(selected_data$Gender == 1, "male", "female"))
selected_data$scholarship_holder <- factor(ifelse(selected_data$scholarship_holder == 1, "yes", "no"))
```

```{r}
skim(selected_data)
```

## Research question

**What are the significant factors influencing first-year academic performance among students, as measured by the average first-year grade?**

It's interesting for us to have idea what we could change to improve our grades (become older for example :3).
But overall understanding the factors that influence academic performance is crucial for improving educational outcomes. This can not only be useful for students to have ideas what to change to get better marks but also can guide universities in designing interventions, such as targeted support for specific groups.
