---
title: "Grade Mark Correlation Features for University Students"
author: "Bananchiki Group: Lev Laptinov, Dmytro Kolisnyk, Vasyl Moiseiev"
---

# Introduction

[Source of dataset](https://archive.ics.uci.edu/dataset/697/predict+students+dropout+and+academic+success)

A dataset created from a higher education institution (acquired from several disjoint databases) related to students enrolled in different undergraduate degrees, such as agronomy, design, education, nursing, journalism, management, social service, and technologies. The dataset includes information known at the time of student enrollment (academic path, demographics, and social-economic factors) and the students' academic performance at the end of the first and second semesters. [Dataset is already performed a rigorous data preprocessing to handle data from anomalies, unexplainable outliers, and missing values.](https://arc.net/l/quote/emkqvmzr)

```{r, include=FALSE}
library(tidyverse)
library(rvest)
library(ggplot2)
if (!require("skimr")) {
  install.packages("skimr")
}
library(skimr)
library(tidymodels)
```

# Data Loading and Preprocessing

```{r, include=FALSE}
data <- read_delim("data.csv", delim = ";")
head(data)
```

```{r}
selected_data <- data %>%
  select(
    attendance_type = `Daytime/evening attendance\t`,
    previous_qualification_grade = `Previous qualification (grade)`,
    admission_grade = `Admission grade`,
    gender = `Gender`,
    scholarship_holder = `Scholarship holder`,
    age_at_enrollment = `Age at enrollment`,
    first_sem_grade = `Curricular units 1st sem (grade)`,
    second_sem_grade = `Curricular units 2nd sem (grade)`
  ) %>%
  filter(
    admission_grade > 0,
    previous_qualification_grade > 0,
    first_sem_grade > 0,
    second_sem_grade > 0
  ) %>%
  mutate(
    first_year_grade = (first_sem_grade + second_sem_grade) / 2,
    attendance_type = factor(if_else(attendance_type == 1, "daytime", "evening")),
    gender = factor(if_else(gender == 1, "male", "female")),
    scholarship_holder = factor(if_else(scholarship_holder == 1, "yes", "no"))
  )
```



# Research Question

**What are the significant factors influencing first-year academic performance among students, as measured by the average first-year grade?**

Understanding the factors that influence academic performance is crucial for improving educational outcomes. This can not only be useful for students to improve their grades but also guide universities in designing interventions, such as targeted support for specific groups.

# Exploratory Data Analysis

```{r}
skim(selected_data)
```

## Distribution of Grades

```{r}
sturges_bins <- function(data) {
  ceiling(1 + log2(length(data)))
}

# First-Year Grade Distribution
first_year_bins <- sturges_bins(selected_data$first_year_grade)
ggplot(selected_data, aes(x = first_year_grade)) +
  geom_histogram(bins = first_year_bins, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(
    title = "Histogram of First-Year Grade",
    x = "First-Year Grade",
    y = "Count"
  )

# First-Semester Grade Distribution
first_sem_bins <- sturges_bins(selected_data$first_sem_grade)
ggplot(selected_data, aes(x = first_sem_grade)) +
  geom_histogram(bins = first_sem_bins, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(
    title = "Histogram of First-Semester Grade",
    x = "First-Semester Grade",
    y = "Count"
  )

# Second-Semester Grade Distribution
second_sem_bins <- sturges_bins(selected_data$second_sem_grade)
ggplot(selected_data, aes(x = second_sem_grade)) +
  geom_histogram(bins = second_sem_bins, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(
    title = "Histogram of Second-Semester Grade",
    x = "Second-Semester Grade",
    y = "Count"
  )

# Previous Qualification Grade Distribution
previous_qualification_bins <- sturges_bins(selected_data$previous_qualification_grade)
ggplot(selected_data, aes(x = previous_qualification_grade)) +
  geom_histogram(bins = previous_qualification_bins, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(
    title = "Histogram of Previous Qualification Grade",
    x = "Previous Qualification Grade",
    y = "Count"
  )
```

## Correlations

```{r}
cor_first_sem <- cor(selected_data$admission_grade, selected_data$first_sem_grade, use = "complete.obs")
cor_second_sem <- cor(selected_data$admission_grade, selected_data$second_sem_grade, use = "complete.obs")
cor_first_year <- cor(selected_data$admission_grade, selected_data$first_year_grade, use = "complete.obs")
cor_previous_qualification <- cor(selected_data$admission_grade, selected_data$previous_qualification_grade, use = "complete.obs")

list(
  cor_first_sem = cor_first_sem,
  cor_second_sem = cor_second_sem,
  cor_first_year = cor_first_year,
  cor_previous_qualification = cor_previous_qualification
)
```

# Linear and Nonlinear Relationships

## Admission Grade vs First-Year Grade

```{r}
# Linear and Nonlinear Models
selected_data <- selected_data %>%
  mutate(admission_grade_squared = admission_grade^2)

linear_fit <- lm(first_year_grade ~ admission_grade, data = selected_data)
nonlinear_fit <- lm(first_year_grade ~ admission_grade + admission_grade_squared, data = selected_data)

# Model Summaries
summary(linear_fit)
summary(nonlinear_fit)

# Visualization
r2_linear <- glance(linear_fit)$r.squared
r2_nonlinear <- glance(nonlinear_fit)$r.squared

list(
  r2_linear = r2_linear,
  r2_nonlinear = r2_nonlinear
)

ggplot(selected_data, aes(x = admission_grade, y = first_year_grade)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = FALSE, color = "red") +
  labs(
    title = "Nonlinear Relationship Between Admission Grade and First-Year Grade",
    x = "Admission Grade",
    y = "First-Year Grade"
  ) +
  theme_minimal()
```

## Previous Qualification Grade vs First-Year Grade

```{r}
# Linear and Nonlinear Models
selected_data <- selected_data %>%
  mutate(previous_qualification_grade_squared = previous_qualification_grade^2)

linear_fit <- lm(first_year_grade ~ previous_qualification_grade, data = selected_data)
nonlinear_fit <- lm(first_year_grade ~ previous_qualification_grade + previous_qualification_grade_squared, data = selected_data)

# Model Summaries
summary(linear_fit)
summary(nonlinear_fit)

# Visualization
r2_linear <- glance(linear_fit)$r.squared
r2_nonlinear <- glance(nonlinear_fit)$r.squared

list(
  r2_linear = r2_linear,
  r2_nonlinear = r2_nonlinear
)

ggplot(selected_data, aes(x = previous_qualification_grade, y = first_year_grade)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = FALSE, color = "red") +
  labs(
    title = "Nonlinear Relationship Between Previous Qualification Grade and First-Year Grade",
    x = "Previous Qualification Grade",
    y = "First-Year Grade"
  ) +
  theme_minimal()
```


```{r}
ggplot(selected_data, aes(x = admission_grade, y = first_year_grade, color = gender, shape = factor(gender))) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_smooth(
    method = "lm", se = FALSE, color = "gray", linetype = "dashed",
    aes(group = 1)
  ) +
  labs(
    title = "First-Year Grade vs Admission Grade with gender Trends",
    x = "Admission Grade",
    y = "First-Year Grade",
    color = "Gender",
    shape = "Gender"
  ) +
  theme_minimal()
```
```{r}
ggplot(selected_data, aes(x = admission_grade, y = first_year_grade, color = scholarship_holder, shape = factor(scholarship_holder))) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_smooth(
    method = "lm", se = FALSE, color = "gray", linetype = "dashed",
    aes(group = 1)
  ) +
  labs(
    title = "First-Year Grade vs Admission Grade with Scholarship Holder Trends",
    x = "Admission Grade",
    y = "First-Year Grade",
    color = "Scholarship Holder",
    shape = "Scholarship Holder"
  ) +
  theme_minimal()
```
```{r}
ggplot(selected_data, aes(x = admission_grade, y = first_year_grade, color = attendance_type, shape = factor(attendance_type))) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_smooth(
    method = "lm", se = FALSE, color = "gray", linetype = "dashed",
    aes(group = 1)
  ) +
  labs(
    title = "First-Year Grade vs Admission Grade with Attendance Type Trends",
    x = "Admission Grade",
    y = "First-Year Grade",
    color = "Attendance Type",
    shape = "Attendance Type"
  ) +
  theme_minimal()
```
