---
title: "Project 2"
author: "Bananchiki group: Lev Laptinov, Dmytro Kolisnyk, Vasyl Moiseiev"
---

# Introduction

A dataset created from a higher education institution (acquired from several disjoint databases) related to students enrolled in different undergraduate degrees, such as agronomy, design, education, nursing, journalism, management, social service, and technologies. The dataset includes information known at the time of student enrollment (academic path, demographics, and social-economic factors) and the students' academic performance at the end of the first and second semesters.

```{r}
library(tidyverse)
library(rvest)
library(ggplot2)
```

```{r}
data <- read_delim("data.csv", delim = ";")
head(data)

```

```{r}
colnames(data)
```

```{r}
selected_data <- data %>%
    select(
        attendance_type = `Daytime/evening attendance\t`,
        previous_qualification_grade = `Previous qualification (grade)`,
        admission_grade = `Admission grade`,
        `Gender`,
        scholarship_holder= `Scholarship holder`,
        age_at_enrollment = `Age at enrollment`,
        first_sem_grade = `Curricular units 1st sem (grade)`,
        second_sem_grade = `Curricular units 2nd sem (grade)`
    )
head(selected_data)
```

```{r}
selected_data$first_year_grade <- (selected_data$first_sem_grade + selected_data$second_sem_grade)/2
selected_data$attendance_type <- ifelse(selected_data$attendance_type == 1, "daytime", "evening")
selected_data$Gender <- ifelse(selected_data$Gender == 1, "male", "female")
selected_data$scholarship_holder <- ifelse(selected_data$scholarship_holder == 1, "yes", "no")
```

```{r}
get_feature_info <- function(x) {
  if (is.character(x)) {
    paste("Unique values:", unique(x) %>% paste(collapse = ", "))
  } else {
    paste("Min:", min(x), "Max:", max(x), "Mean:", mean(x), "Median:", median(x), "SD:", sd(x))
}
}

variable_info <- data.frame(
  Variable = names(selected_data),
  DataType = sapply(selected_data, class),
  Feature_info = sapply(selected_data, get_feature_info)
)
variable_info
```

## Research question

**Which featrues have significant with result grade for the first semester?**

It's interesting for us to have idea what we could change to improve our grades (become older for example :3).