---
title: "Grade Mark Correlation Features for University Students"
author: "Lev Laptinov, Dmytro Kolisnyk, Vasyl Moiseiev"
toc: true
toc-depth: 2
---

## Introduction

[Source of dataset](https://archive.ics.uci.edu/dataset/697/predict+students+dropout+and+academic+success)

A dataset created from a higher education institution (acquired from several disjoint databases) related to students enrolled in different undergraduate degrees, such as agronomy, design, education, nursing, journalism, management, social service, and technologies. The dataset includes information known at the time of student enrollment (academic path, demographics, and social-economic factors) and the students' academic performance at the end of the first and second semesters. Dataset is already performed a rigorous data preprocessing to handle data from anomalies, unexplainable outliers, and missing values.

```{r, include=FALSE}
library(tidymodels) 
library(tidyverse)
library(skimr) # For summarizing and understanding data, providing detailed overviews (e.g., mean, missing values).
library(knitr) # For creating elegant tables and dynamic reports, especially in RMarkdown documents.
library(patchwork) # To combine multiple ggplot2 plots into a single cohesive layout (e.g., grids or stacked plots).
```

## Data Loading, Filtering and Selection
```{r, include=FALSE}
data <- read_delim("data.csv", delim = ";")
head(data)
```

1. **Selecting Relevant Columns**:
   - Columns are renamed and selected using `select()`, making the dataset easier to work with.
   - Example:
     - `Daytime/evening attendance\t` is renamed to `attendance_type`.
     - `Previous qualification (grade)` is renamed to `previous_qualification_grade`.
   - This process ensures that only the necessary columns for analysis are retained.

2. **Filtering Rows**:
   - Rows with grades (`admission_grade`, `previous_qualification_grade`, `first_sem_grade`, `second_sem_grade`) greater than 0 are retained using `filter()`.
   - This step removes rows with invalid or missing grade values.

### Data Transformation
3. **Adding a New Column**:
   - `first_year_grade`: The average of `first_sem_grade` and `second_sem_grade`, calculated as `(first_sem_grade + second_sem_grade) / 2`. This summarizes the student's overall performance in the first year.

4. **Transforming Columns**:
   - **`attendance_type`**:
     - Values are transformed into a factor where `1` is labeled "daytime" and other values are labeled "evening."
   - **`gender`**:
     - Values are transformed into a factor where `1` is labeled "male" and other values are labeled "female."
   - **`scholarship_holder`**:
     - Values are transformed into a factor where `1` is labeled "yes" and other values are labeled "no."
   - These transformations standardize the data and make categorical variables more interpretable.

```{r}
selected_data <- data %>%
  select(
    attendance_type = `Daytime/evening attendance\t`,
    previous_qualification_grade = `Previous qualification (grade)`,
    admission_grade = `Admission grade`,
    gender = `Gender`,
    scholarship_holder = `Scholarship holder`,
    age_at_enrollment = `Age at enrollment`,
    first_sem_grade = `Curricular units 1st sem (grade)`,
    second_sem_grade = `Curricular units 2nd sem (grade)`
  ) %>%
  filter(
    admission_grade > 0,
    previous_qualification_grade > 0,
    first_sem_grade > 0,
    second_sem_grade > 0
  ) %>%
  mutate(
    first_year_grade = (first_sem_grade + second_sem_grade) / 2,
    attendance_type = factor(if_else(attendance_type == 1, "daytime", "evening")),
    gender = factor(if_else(gender == 1, "male", "female")),
    scholarship_holder = factor(if_else(scholarship_holder == 1, "yes", "no"))
  )
```



## Research Question

**How do admission grades, previous qualification grades, and subgroup characteristics (gender, scholarship status, attendance type) influence first-year academic performance among university students, as measured by the average first-year grade?**

Understanding the factors that influence academic performance is crucial for improving educational outcomes. This can not only be useful for students to improve their grades but also guide universities in designing interventions, such as targeted support for specific groups.

## Exploratory Data Analysis

```{r}
skim(selected_data)
```

## Distribution of Grades

```{r}
sturges_bins <- function(data) {
  ceiling(1 + log2(length(data)))
}


# Calculate bins for each histogram
admission_bins <- sturges_bins(selected_data$admission_grade)
first_sem_bins <- sturges_bins(selected_data$first_sem_grade)
second_sem_bins <- sturges_bins(selected_data$second_sem_grade)
previous_qualification_bins <- sturges_bins(selected_data$previous_qualification_grade)

# Create individual plots
admission_plot <- ggplot(selected_data, aes(x = admission_grade)) +
  geom_histogram(bins = admission_bins, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(
    title = "Admission Grade",
    x = "Admission Grade",
    y = "Count"
  )

first_sem_plot <- ggplot(selected_data, aes(x = first_sem_grade)) +
  geom_histogram(bins = first_sem_bins, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(
    title = "First-Semester Grade",
    x = "First-Semester Grade",
    y = "Count"
  )

second_sem_plot <- ggplot(selected_data, aes(x = second_sem_grade)) +
  geom_histogram(bins = second_sem_bins, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(
    title = "Second-Semester Grade",
    x = "Second-Semester Grade",
    y = "Count"
  )

previous_qualification_plot <- ggplot(selected_data, aes(x = previous_qualification_grade)) +
  geom_histogram(bins = previous_qualification_bins, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(
    title = "Previous Qualification Grade",
    x = "Previous Qualification Grade",
    y = "Count"
  )

# Combine all plots into a grid
(admission_plot | first_sem_plot) / (second_sem_plot | previous_qualification_plot)
```

Based on the provided data summary and histogram visualizations:

1. **Distributions**:
   - The numeric variables, such as "Admission Grade," "First-Semester Grade," "Second-Semester Grade," and "Previous Qualification Grade," appear to follow approximately normal distributions with some variations in the tails.

2. **Student Characteristics**:
   - The dataset includes 3,512 rows and nine columns, with three categorical variables and six numeric ones.
   - The majority of students (3,147) are enrolled in day attendance, while a smaller group (365) are in evening attendance.
   - There are more female students (2,445) than male students (1,067).

3. **Scholarship Holders**:
   - About 28.6% (1,003 out of 3,512) of the students hold scholarships, while the majority (2,509) do not.

4. **Grades**:
   - The "Previous Qualification Grade" has a higher mean (132.63) and a larger spread compared to "Admission Grade" and semester grades, which cluster around similar mean values (approximately 127.14 and 12.75, respectively).

## Correlations

```{r}
cor_first_sem <- cor(selected_data$admission_grade, selected_data$first_sem_grade, use = "complete.obs")
cor_second_sem <- cor(selected_data$admission_grade, selected_data$second_sem_grade, use = "complete.obs")
cor_previous_qualification <- cor(selected_data$admission_grade, selected_data$previous_qualification_grade, use = "complete.obs")

list(
  cor_first_sem = cor_first_sem,
  cor_second_sem = cor_second_sem,
  cor_previous_qualification = cor_previous_qualification
)
```
### Hypothesis
**Null Hypothesis (H0):** There is no significant correlation between admission grades and academic performance (first-semester grade, second-semester grade, and previous qualification grade).  
**Alternative Hypothesis (H1):** Admission grades are significantly correlated with academic performance (first-semester grade, second-semester grade, and previous qualification grade).

---

### Methodology
Calculate the Pearson correlation coefficient between "Admission Grade" and the following:
  - First-Semester Grade
  - Second-Semester Grade
  - Previous Qualification Grade


### Results and Interpretation
- **First-Semester Grade vs. Admission Grade**:
  - Correlation: $r = 0.31$
  - Interpretation: A low to moderate positive correlation exists, indicating that admission grades have some predictive power for first-semester performance.

- **Second-Semester Grade vs. Admission Grade**:
  - Correlation: $r = 0.28$
  - Interpretation: A low positive correlation, suggesting admission grades are less predictive for second-semester performance compared to the first semester.

- **Previous Qualification Grade vs. Admission Grade**:
  - Correlation: $r = 0.60$
  - Interpretation: A moderate positive correlation, indicating a stronger relationship between previous qualifications and admission grades.

### Conclusion
- There is evidence to support a correlation between admission grades and academic performance metrics. 
- The relationship is strongest between "Admission Grade" and "Previous Qualification Grade" ($r = 0.60$), suggesting that past academic performance is one of a key factor in determining admission grades.
- The weaker correlations with first and second-semester grades imply that other factors beyond admission grades influence academic success during the semesters.
- Future research could involve regression modeling to explore additional predictors of academic performance.


## Linear Relationships


### Methodology

1. **Objective**:
   - Analyze the relationships between admission grades and first-year grades as well as previous qualification grades.
   - Fit linear and quadratic regression models to evaluate which relationship better explains the variation.

2. **Steps**:
   - **Data Preparation**:
     - Transform the data by creating a squared term for admission grades.
   - **Model Fitting**:
     - Fit a linear regression model and a quadratic regression model for:
       - Admission Grade vs. First-Year Grade.
       - Admission Grade vs. Previous Qualification Grade.
   - **Model Evaluation**:
     - Extract $R^2$ values for each model to determine the goodness of fit.
     - Compare the linear and quadratic models for each pair of variables.
   - **Visualization**:
     - Plot scatter plots with regression lines to visually assess trends.

3. **Threshold for Interpretation**:
   - A higher $R^2$ value indicates a better fit.
   - Statistical significance is assessed using p-values for model terms ($p < 0.05$ is significant).


### Admission Grade vs First-Year Grade

```{r}
# Linear and squared Models
selected_data <- selected_data %>%
  mutate(admission_grade_squared = admission_grade^2)

linear_fit <- lm(first_year_grade ~ admission_grade, data = selected_data)
squared_fit <- lm(first_year_grade ~ admission_grade_squared, data = selected_data)

# Model Summaries
kable(tidy(linear_fit))
kable(tidy(squared_fit))

# Visualization
r2_linear <- glance(linear_fit)$r.squared
r2_squared <- glance(squared_fit)$r.squared

list(
  r2_linear = r2_linear,
  r2_squared = r2_squared
)

ggplot(selected_data, aes(x = admission_grade, y = first_year_grade)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(
    title = "Relationship Between Admission Grade and First-Year Grade",
    x = "Admission Grade",
    y = "First-Year Grade"
  ) +
  theme_minimal()
```

### Admission Grade vs Previous Qualification Grade

```{r}
linear_fit <- lm(previous_qualification_grade ~ admission_grade, data = selected_data)
squared_fit <- lm(previous_qualification_grade ~ admission_grade_squared, data = selected_data)

# Model Summaries
kable(tidy(linear_fit))
kable(tidy(squared_fit))

# Visualization
r2_linear <- glance(linear_fit)$r.squared
r2_squared <- glance(squared_fit)$r.squared

list(
  r2_linear = r2_linear,
  r2_squared = r2_squared
)

ggplot(selected_data, aes(x = admission_grade, y = previous_qualification_grade)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(
    title = "Relationship Between Admission Grade and Previous Qualification Grade",
    x = "Admission Grade",
    y = "Previous Qualification Grade",
  ) +
  theme_minimal()
```

### Results and Conclusion

#### 1. Admission Grade vs. First-Year Grade
- **Linear Model**:
  - Slope: 0.0285 (p-value < 0.001), significant.
  - $R^2$: 0.1026, indicating only 10.26% of the variability in first-year grades is explained by admission grades.
- **Quadratic Model**:
  - Coefficient for squared term: 0.0001 (p-value < 0.001), significant.
  - $R^2$: 0.1040, only marginally better than the linear model.
- **Conclusion**:
  - The relationship is weak, with both linear and quadratic models explaining limited variation. Other factors likely influence first-year grades beyond admission grades.

#### 2. Admission Grade vs. Previous Qualification Grade
- **Linear Model**:
  - Slope: 0.5604 (p-value < 0.001), significant.
  - $R^2$: 0.3558, indicating that 35.58% of the variability in previous qualification grades is explained by admission grades.
- **Quadratic Model**:
  - Coefficient for squared term: 0.0021 (p-value < 0.001), significant.
  - $R^2$: 0.3580, slightly better than the linear model.
- **Conclusion**:
  - The relationship is moderately strong, and the quadratic model explains the variation slightly better, but not enough to suggest something other than linear trend.

## Investigation of Subgroups

### Methodology

1. **Objective**:
   - Investigate how the relationship between admission grades and first-year grades varies by gender, scholarship status, and attendance type.

2. **Analysis Steps**:
   - **Data Visualization**:
     - Scatterplots were generated for each subgroup (gender, scholarship holder, attendance type) with distinct colors and shapes.
     - Separate linear regression trends were overlaid for each subgroup and a combined dashed trend for overall comparison.
   - **Modeling**:
     - Separate linear regression models for each subgroup to assess differences in slopes and intercepts.
   - **Comparison**:
     - Visualize and compare the trends for subgroups to identify patterns and deviations from the overall trend.

3. **Tools**:
   - **Visualization**:
     - Used `ggplot2` for creating scatterplots with subgroup-specific and overall regression trends.
   - **Interpretation**:
     - Focused on how subgroup-specific trends deviate from the overall trend, emphasizing slope differences and clustering of data points.

```{r}
ggplot(selected_data, aes(x = admission_grade, y = first_year_grade, color = gender, shape = factor(gender))) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_smooth(
    method = "lm", se = FALSE, color = "gray", linetype = "dashed",
    aes(group = 1)
  ) +
  labs(
    title = "First-Year Grade vs Admission Grade with gender Trends",
    x = "Admission Grade",
    y = "First-Year Grade",
    color = "Gender",
    shape = "Gender"
  ) +
  theme_minimal()
```
```{r}
ggplot(selected_data, aes(x = admission_grade, y = first_year_grade, color = scholarship_holder, shape = factor(scholarship_holder))) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_smooth(
    method = "lm", se = FALSE, color = "gray", linetype = "dashed",
    aes(group = 1)
  ) +
  labs(
    title = "First-Year Grade vs Admission Grade with Scholarship Holder Trends",
    x = "Admission Grade",
    y = "First-Year Grade",
    color = "Scholarship Holder",
    shape = "Scholarship Holder"
  ) +
  theme_minimal()
```
```{r}
ggplot(selected_data, aes(x = admission_grade, y = first_year_grade, color = attendance_type, shape = factor(attendance_type))) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_smooth(
    method = "lm", se = FALSE, color = "gray", linetype = "dashed",
    aes(group = 1)
  ) +
  labs(
    title = "First-Year Grade vs Admission Grade with Attendance Type Trends",
    x = "Admission Grade",
    y = "First-Year Grade",
    color = "Attendance Type",
    shape = "Attendance Type"
  ) +
  theme_minimal()
```

### Conclusions

#### Gender
- **Observation**:
  - Both male and female trends exhibit a positive relationship between admission grades and first-year grades.
  - The trendline for females has a slightly higher slope, suggesting that admission grades may better predict first-year performance for females compared to males.
- **Implication**:
  - Gender differences in performance predictors might reflect different academic strengths or external factors affecting grades.

#### Scholarship Holders
- **Observation**:
  - Scholarship holders show a higher trendline compared to non-scholarship holders.
  - This suggests that students with scholarships tend to have better first-year grades at similar levels of admission grades.
- **Implication**:
  - Scholarship status may capture additional factors, such as motivation or prior achievements, influencing first-year success.

#### Attendance Type
- **Observation**:
  - Daytime students exhibit a steeper trend compared to evening students.
  - Evening students have generally lower first-year grades for equivalent admission grades.
- **Implication**:
  - Attendance type might be influenced by external commitments (e.g., work for evening students) that impact academic performance.

## Results and Discussion

This analysis has explored the relationships between admission grades, previous qualification grades, and first-year academic performance for university students. The findings reveal:

1. **Admission Grades as Predictors**:
   - Admission grades are moderately correlated with previous qualification grades ($r = 0.60$) but show weaker correlations with first-semester ($r = 0.31$) and second-semester ($r = 0.28$) grades. This suggests that while admission grades are influenced by prior academic achievements, they are less predictive of subsequent university performance.

2. **Linear Relationships**:
    - **Admission Grades vs. First-Year Grades**:
      - The relationship is weak, with admission grades explaining only 10.26% of the variability in first-year grades.
      - Both linear and quadratic models show limited predictive power, indicating that other factors likely influence first-year academic performance.

    - **Admission Grades vs. Previous Qualification Grades**:
      - Admission grades moderately predict previous qualification grades, explaining 35.58% of the variability.
      - The relationship remains primarily linear, with only a slight improvement in prediction when using a quadratic model.

3. **Impact of Subgroup Characteristics**:
   Gender, scholarship status, and attendance type impact the relationship between admission grades and first-year performance:
     - Female students and scholarship holders display stronger positive relationships between admission grades and academic outcomes.
     - Scholarship holders show a higher trendline compared to non-scholarship holders
     - Daytime students outperform evening students for similar admission grades.

### Appendix

Lev Laptinov - Introduciton, Data Loading and Preprocessing, Research Question, Exploratory Data Analysis

Dmytro Kolisnyk - Methodology, Correlation, part of Linear Relationships, Investigation of Subgroups

Vasyl Moiseiev - part of Linear Relationships, Discussion and Conclusion, presentation